Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach

### 2014-05-23 manuscripts/manuscript.Rmd
updates as sent to steve

To construct the value function, we consider a return when $X_t=x$ and harvest $h_t=h$ denoted by
$\Pi(x,h)$. For example, if the return is simply the harvest at time $t$, then $\Pi(x,h)=min(x,h)$. We assume that future harvests are discounted relative to current ones at a constant rate of discount $\delta$ and ask for the harvest policy that maximizes total discounted harvest between the current time $t$ and a final time $T$, that is we seek to maximize over choices of harvest $\mathbf{E_Z}[ \sum_{t = 0}^{T}  \Pi_t(X_t, h_t) \delta^t]$, where the state dynamics are given by Eqn 1 and $\mathbf{E_Z}$ denotes the expectation over the process stochasticity.

In order to find that policy, we introduce the value function$V_t(x_t)$ representing the total discounted catch from time $t$ onwards given that $X_t=x$.  This value function satisfies an equation of Stochastic Dynamic Programming  [SDP, @Mangel1988; @Clark2000, @Clark2009], sometimes called the Bellman-Hamilton-Jacobi equation [@Mangel2014].

\begin{equation}
V_t(x) = \max_{h_t} \left[\Pi(h_t, x_t) + \delta \cdot \mathbf{E_Z}  V_{t+1}( Z_{t+1} f(x - h_t)) \right]
\end{equation}

That is, at time $t$, when population size is $x$ and harvest $h_t$ is applied, the immediate return is $\Pi(h_t, x_t)$ and the population size after the take is $x-h_t$, which is then translated into $X_{t+1}$ by Eqn 1. We consider the finite time problem with $T=$`r Tmax`, which we solve using value iteration algorithm [@Mangel1988; @Clark2000].

@@

@@

The nonparametric moniker for the
GP approach is unfortunate as (a) it still involves the estimation of
parameters, and (b), statisticians also use non-parametric to mean models
that make only weak distributional assumptions  [e.g. @Lehmann1975].
However, we retain the terminology for consistency with previous studies.

<!-- FIXME Marc suggests putting the full GP formulation in here?? -->

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a SDP algorithm.  Recall that unlike
the parametric models the Gaussian process with fixed parameters already
predicts a distribution of curves rather than a single curve. We must
first integrate over this distribution of curves given a sampling of
parameter values drawn from the posterior distribution of the two GP
parameters, before integrating over the posterior of those parameters
themselves.

@@

Large noise relative to the distance between the stable and unstable
point, (or more precisely, between the optimal escapement and the unstable
point) increases the chance of a stochastic transition.  As a result,
the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be modeled
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.

@@

The degree to which the decision-making part of the algorithm (the SDP)
chooses to explore or avoid the resulting region of uncertainty can also
be influenced by the curvature of the value (profit) function $\Pi$. Both
to simplify the intuition and avoid biasing this result, we have chosen
profits that are linear in the catch and thus neither risk-prone nor
risk adverse.  Making this function concave, representing the typical
assumption of diminishing returns, would make the SDP more risk-adverse
(as larger-than-expected stock sizes offer diminished returns relative
to the cost of smaller-than-expected stock sizes), and strengthen the
result shown here in which the BNP solution tends to avoid the region
of uncertainty.  Sufficiently convex or risk-prone functions could lead
the SDP to attempt higher exploitation rates despite the uncertainty.
Understanding the relative roles of such functions would be a promising
direction for future investigation.

@@

Gaussian process model
----------------------

Assuming the data are known with some process noise,

$\varepsilon$ IID normal, variance $\sigma_n^2$. Then under the GP,

$$f(x) \sim \mathcal{N}(E,C)$$
$$E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y$$
$$C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)$$

Where $K$ is the Gaussian process kernel.  Throughout, we use the radial basis function kernel:

The marginal likelihood is then given by:

$$\log(p(y | X)) = -\tfrac{1}{2} \mathbf{y}^T (K + \sigma_n^2 \mathbf{I})^{-1} y  - \tfrac{1}{2} \log\left| K + \sigma_n^2 \mathbf{I} \right| - \tfrac{n}{2}\log 2 \pi$$

See @Munch2005a or @Rasmussen2006 for a more detailed introduction.

@@

The Gaussian process priors on both the lengthscale $\ell$ and process noise $\sigma$ parameters are both vague inverse Gamma,

@@

Optimal Control Problem Specification 
-------------------------------------

We seek the harvest policy $h(x)$ that maximizes:

$$ \max_{h_t} \sum_{t \in 0}^{\infty}  \Pi_t(X_t, h_t) \delta^t  $$

subject to the profit function $\Pi(X_t,h)$, discount rate $\delta$, and the state
equation

$$X_{t+1} = Z_t f(S_t)  $$
$$S_t = X_t - h_t $$

For simplicity we assume profit function $\Pi_t(X_t, h_t)$ is linear in the realized harvest, only
enforcing the restriction that harvest can not exceed available stock, that is $\Pi(h,x) = min(h,x)$. $Z_t$ is as previously defined.

From this we can write down the Bellman recursion as:

$$V_t(x_t) = \max_h \mathbf{E} \left(\Pi(h_t, x_t) + \delta V_{t+1}( Z_{t+1} f(x_t - h_t)) \right)$$

Where $V_t$ is the discounted value of the profit given that $X(t) = x$

>>>>

To construct the value function, we consider a return when $X_t=x$ and harvest $h_t=h$ denoted as the reward, $R(x,h)$. For example, if the return is simply the harvest at time $t$, then $R(x,h)=min(x,h)$. We assume that future harvests are discounted relative to current ones at a constant rate of discount $\delta$ and ask for the harvest policy that maximizes total discounted harvest between the current time $t$ and a final time $T$, that is we seek to maximize over choices of harvest $\mathbf{E}_{X_t+1} [ \sum_{t = 0}^{T}  R_t(X_t, h_t) \delta^t]$, where the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the expectation over the process stochasticity in the future population state.

In order to find that policy, we introduce the value function$V_t(x_t)$ representing the total discounted catch from time $t$ onwards given that $X_t=x$.  This value function satisfies an equation of Stochastic Dynamic Programming  [SDP, @Mangel1988; @Clark2000; @Clark2009], sometimes called the Bellman-Hamilton-Jacobi equation [@Mangel2014].

\begin{equation}
V_t(x) = \max_{h_t} \left[R(h_t, x_t) + \delta \cdot \mathbf{E}_{X_t+1} V_{t+1}( Z_{t+1} f(x - h_t)) \right]
\end{equation}

That is, at time $t$, when population size is $x$ and harvest $h_t$ is applied, the immediate return is $R(h_t, x_t)$ and the population size after the take is $x-h_t$, which is then translated into $X_{t+1}$ by Eqn 1. We consider the finite time problem with $T=$`r MaxT`, which we solve using value iteration algorithm [@Mangel1988; @Clark2000].

@@

The core difference for our purpose between the GP representation and the parametric
models we have already introduced is the "non-parametric" property, which
means that an estimated GP model is defined explicitly in reference to 
some observed data.  Unlike the models above, it cannot be specified by
the value of some parameters alone.  The nonparametric moniker for the
GP approach is unfortunate as (a) it still involves the estimation of
parameters, and (b), statisticians also use non-parametric to mean models
that make only weak distributional assumptions  [e.g. @Lehmann1975].
However, we retain the terminology for consistency with previous studies.

@@

A GP is an infinite dimensional (because it considers a continuous random
variable) stochastic process for which any realization is a multivariate
normal distribution. Thus, to characterize the GP we need a mean function
and a covariance matrix. We proceed as follows.

Assuming the data $X_o$ are observed with some process noise,

$\varepsilon$ IID normal, variance $\sigma_n^2$. Then under the GP, the predicted data $X_p$ given the observations obeys

$$f(X_p|X_o) \sim \mathcal{N}(E,C)$$
$$E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y$$
$$C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)$$

Where $K$ is the Gaussian process kernel.  (Compare to  Eqn 2.22 of @Rasmussen2006 for a more detailed derivation.) Throughout, we use the radial basis function kernel:

We use vague inverse Gamma priors on both the lengthscale $\ell$ and process noise $\sigma$ parameters, of the form

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a SDP algorithm. The only difference 
is that the uncertainty in the future state under the GP, $X_{t+1}$,
includes both process uncertainty (based on the estimation of $\sigma$) 
and "structural uncertainty" of the posterior collection of curves.  
(Recall that given the parameters, a parametric model is represented
by a single curve, while the GP remains a distribution of curves).  
From the GP posteriors, we can write down the (discritized) transition
matrix representing the probability of going to each state $X_{t+1}$ 
given any current state $X_t$ and any harvest $h_t$ (See the function
`gp_transition_matrix()` in the provided R package). Given this transition
matrix, we use the same value iteration algorithm as in the parametric
case to determine the optimal policy.

@@

Changing the noise or the distance between stable and unstable points
does not impact the performance of the GP relative to the optimal solution
given the true model and true parameters.  The parametric models are
more senstive to this difference.  Large noise relative to the distance between the stable and unstable
point increases the chance of a stochastic transition.  More precisely,
if we let $L = K - x_c$, then the probability of a transition in a given
window of time $T$ scales as

$$P(x_t < x_c | t \in T) \propto  \exp\left(\frac{L^2}{\sigma^2}\right),$$

(following the Ahrenius relationship, see @Gardiner2009 for the derivation).
As a result, the impact of using a model which underestimates the risk of harvesting
past the critical point is much more severe, since this such a situation
occurs more often.  Conversely, with large enough distance between the
optimal escapement and unstable points relative to the noise, the chance
of a transition becomes vanishingly small and all models can be modeled
near-optimally. Models that underestimate the cost incurred by stock
sizes fluctuating significantly below the optimal escapement level will
not perform poorly as long as those fluctuations are sufficently rare.

@@

The degree to which the decision-making part of the algorithm (the SDP)
chooses to explore or avoid the resulting region of uncertainty can also
be influenced by the curvature of the value (profit) function $R$. Both
to simplify the intuition and avoid biasing this result, we have chosen
profits that are linear in the catch and thus neither risk-prone nor
risk adverse.  Making this function concave, representing the typical
assumption of diminishing returns, would make the SDP more risk-adverse
(as larger-than-expected stock sizes offer diminished returns relative
to the cost of smaller-than-expected stock sizes), and strengthen the
result shown here in which the BNP solution tends to avoid the region
of uncertainty.  Sufficiently convex or risk-prone functions could lead
the SDP to attempt higher exploitation rates despite the uncertainty.
Understanding the relative roles of such functions would be a promising
direction for future investigation.

@@

@@

@@
